---
title: Working with Date and Time
author: Tianyi Shi
date: '2020-03-19'
slug: working-with-date-and-time
categories:
  - sql
tags:
  - sql
  - r
  - datetime
---

```{r setup, include=FALSE}
library(lubridate)
  con <- DBI::dbConnect(RPostgres::Postgres(), host="localhost", dbname="postgres", user='postgres', password='123qazpl,')
```

In R, the package `lubridate` handles date and times. 

# Basic Functions

## Current Date/Time

Current time:

```{sql connection=con}
SELECT NOW();
```

Current date:

```{r}
today()
as_date(now())
```

```{sql connection=con}
SELECT current_time;
```

```{sql connection=con}
SELECT NOW()::DATE;
```

# Addition and Subtraction

In R, `duration`s are always stored in seconds. This can lead to unexpected results:

```{r}
ymd(20000101) + dyears(1)
```

More often `period`s are used:

```{r}
ymd(20000101) + years(1)
```

```{r}
ymd(20000101) + period(year=1, month=1, day=1, hour=1, minute=5, second=15)
```


In SQL, the behaviour is similar to R's `period`s:

```{sql connection=con}
SELECT DATE '2000-01-01' + INTERVAL '1 YEAR 1 MONTH 1 DAY 1 HOUR 5 MINUTE 15 SECOND' AS new_date;
```

Note that this is a datetime. To cast it to date:

```{sql connection=con}
SELECT (DATE '2000-01-01' + INTERVAL '1 YEAR')::DATE;
```

## Time difference (Period/Interval)

```{r}
span <- ymd(20000101) %--% ymd(20010101)
span
as.period(span)
span / ddays(1)
```

This is similar to direct subtraction in 

```{sql connection=con}
SELECT DATE '2001-01-01' - DATE '2000-01-01' AS span
```

The `AGE` function in PostgreSQL is similar to the `%--%` operator followed by `as.period` in lubridate.

```{sql connection=con}
SELECT AGE(DATE '2001-01-01', DATE '2000-01-01') AS span
```

# Extracting Components

In `lubridate`, you use functions such as `year` and `month` on date/datetime to extract components. In PostgreSQL, you use the `EXTRACT` function with this pattern: `EXTRACT(${COMPONENT} FROM date)` where `${COMPONENT}` can be one of those listed in the following table:

|         lubridate          |         PostgreSQL         |     Note     |
| :------------------------: | :------------------------: | :----------: |
|           `year`           |           `YEAR`           |              |
|          `month`           |          `MONTH`           |              |
|      `day` or `mday`       |           `DAY`            | day of month |
|           `wday`           |           `DOW`            | day of week  |
|           `yday`           |           `DOY`            | day of year  |
|             -              |         `CENTURY`          |              |
| `hour`, `minute`, `second` | `HOUR`, `MINUTE`, `SECOND` |              |
|             -              |       `MILLISECOND`        |              |

```{r}
month(now())
```

```{sql connection=con}
SELECT EXTRACT(MILLISECOND FROM NOW());
```




In PostgreSQL, `DOW` (day of week) starts from Sunday.
